---
title: "Bayesian Evaluation of a Mixture Rasch Model Using the bleval Package"
author: "Jieyuan Dong"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Introduction

This document demonstrates the evaluation of a Bayesian mixture Rasch Model using the `bleval` package. The model includes a latent ability (writing knowledge) and a latent class variable. We illustrate the workflow from data preparation to model evaluation, including parameter estimation in CmdStan, convergence diagnostics, and computation of information criteria (DIC, WAIC, LOOIC) and marginal likelihood using the `bleval` package.

## Read Data

We begin by loading the `bleval` package and the dataset `example3_MixRasch`, which contains the observed responses for the mixture Rasch Model.

```{r}
# Install the bleval package
devtools::install_github("luoxh3/bleval")
# Load the package
library(bleval)

# Load the dataset from the bleval package
data(example3_MixRasch, package = "bleval")

D <- 1  # number of dimensions
K <- 2  # number of latent classes
I <- nrow(example3_MixRasch)  # number of respondents (i.e. units)
J <- 7  # number of items

res <- as.matrix(example3_MixRasch)
```

## Parameter Estimation in CmdStan

We define the Bayesian mixture Rasch model in CmdStan and estimate parameters using MCMC sampling.

```{r}
library(cmdstanr)
library(posterior)

stan_code <- "
data {
  int<lower=1> I;                      // number of examminees
  int<lower=1> J;                      // number of items
  int<lower=1> N;                      // number of responses
  int<lower=2> K;                      // number of classes
  array[N] int<lower=1,upper=I> ii;    // person id
  array[N] int<lower=1,upper=J> jj;    // item id
  array[N] int<lower=0,upper=1> y;     // responses
}
parameters {
  positive_ordered[K] lambda;          // ordered component
  array[J] vector[K] beta;		         // item difficulty
  vector<lower=0>[K] sigma;            // sd of person ability within each class
  vector[I] theta;                     // person ability
}
transformed parameters{
  simplex[K] pi = lambda/sum(lambda);	 // mixing proportions
  // Note that pi is the real model parameters, and lambda is introduced and 
  // then normalized to generate pi with an ordered dirichlet prior
}
model {
  vector[K] lpth;
  vector[K] lps; 
  
  // prior
  lambda ~ gamma(0.5, 1);
  sigma ~ normal(0, 5);
  for (j in 1:J)
    beta[j] ~ normal(0, 1);
    
  for (i in 1:I){ 
    for (k in 1:K){ 
      lpth[k] = log(pi[k]) + normal_lpdf(theta[i] | 0, sigma[k]);
    }
    target += log_sum_exp(lpth);
  }

  // likelihood
  for (n in 1:N){
    for (k in 1:K){
      lps[k] = log(pi[k]) + bernoulli_logit_lpmf(y[n] | theta[ii[n]] - beta[jj[n],k]);
    }
    target += log_sum_exp(lps);
  }
}"

data_list <- list(
  K = K,
  I = I,
  J = J,
  N = I*J,
  ii = rep(1:I, each = J),
  jj = rep(1:J, times = I),
  y = as.vector(t(res))
)

# Initialize the CmdStan model
model_file <- write_stan_file(stan_code,  dir = getwd(), basename = "MixRasch")
stan_model <- cmdstan_model(model_file)

# Run the No-U-Turn Sampler with a burn-in period of 3500 iterations
fit <- stan_model$sample(
  data = data_list,
  seed = 1,
  refresh = 500,          
  chains = 4, 
  parallel_chains = 4,
  iter_warmup = 3500,
  iter_sampling = 2500,
  thin = 1,
  show_messages = TRUE
)

# Draw posterior samples from the model
draws <- fit$draws(format="df")

# Extract posterior samples for model parameters only (excluding latent variables)
par_cols <- grep("^beta|^sigma|^lambda", names(draws), value = TRUE)
samples <- as.matrix(draws[, par_cols])

# Extract posterior samples for latent variables
lv_draws <- as.matrix(draws[, grep("^theta", colnames(draws))])
```

## Model Convergence Check

We check the convergence of the MCMC chains using the Gelman-Rubin diagnostic (R-hat). A common threshold for convergence is R-hat \< 1.1.

```{r}
# Summarize the posterior samples and compute the Gelman-Rubin diagnostic (R-hat)
summary <- fit$summary(NULL,
                       posterior::default_summary_measures()[1:4],
                       quantiles = ~ quantile2(., probs = c(0.025, 0.975)),
                       posterior::default_convergence_measures())
# Check the maximum R-hat
max(summary[,"rhat"], na.rm = TRUE)
```

## Model Evaluation with `bleval`

### Compute Information Criteria

**Step 1: Specify the `log_joint_i` function**

The `log_joint_i` function calculates the log joint density for each unit.

```{r}
log_joint_i <- function(samples_s, data, i, Ngrid, nodes) {
  
  # conditional likelihood
  J <- data[["J"]]
  y_i <- data[["y"]][data[["ii"]]==i]
  jj_i <- data.frame(jj=data[["jj"]][data[["ii"]]==i])
  
  lambda1 <- samples_s[1]
  lambda2 <- samples_s[2]
  beta1 <- samples_s[3:(J+2)]
  beta2 <- samples_s[(J+3):(2*J+2)]
  sigma1 <- samples_s[2*J+3]
  sigma2 <- samples_s[2*J+4]
  
  pi1 <- lambda1/(lambda1 + lambda2)
  pi2 <- lambda2/(lambda1 + lambda2)
  
  log_con_i_j <- apply(jj_i,1,function(j){
    log(pi1*dbinom(y_i[j], 1, 1/(1+exp(-(nodes - beta1[j]))), log = FALSE) +
      pi2*dbinom(y_i[j], 1, 1/(1+exp(-(nodes - beta2[j]))), log = FALSE))
  })
  
  if(is.vector(log_con_i_j))
    log_con_i <- sum(log_con_i_j) else 
    log_con_i <- rowSums(log_con_i_j)
    
  # log prior density of latent variables
  log_lv_i <- log(pi1*dnorm(nodes, mean=0, sd=sigma1, log = FALSE)+
    pi2*dnorm(nodes, mean=0, sd=sigma2, log = FALSE))
  
  # return
  log_lv_i + log_con_i
}
```

**Step 2: Compute the posterior means and covariance matrices of latent variables**

```{r}
# Create lists to store posterior mean and variance of latent ability
lv_list <- list()
for (i in 1:I){
  lv_list[[i]] <- lv_draws[ , ((i - 1)*D + 1):(i*D)]
}

# Compute the posterior mean of the latent ability for unit i
lv_mu_list <- lapply(lv_list, mean)
# Compute the posterior variance of the latent ability for unit i
lv_cov_list <- lapply(lv_list, var)

# Make sure that N denotes the number of units (i.e. respondents) in the data to be passed to the functions below
data_list$N <- data_list$I
```

**Step 3: Compute the log-likelihood with latent variables integrated out**

The `log_lik` function from the `bleval` package computes the log likelihood by integrating out the latent variables using adaptive Gauss-Hermite quadrature.

```{r}
log_lik_result <- bleval::log_lik(samples = samples, data = data_list, Ngrid = 5,
                                  lv_mu = lv_mu_list, lv_cov = lv_cov_list,
                                  log_joint_i = log_joint_i, n_cores = 10)
```

**Step 4: Compute information criteria**

The `calc_IC` function from the `bleval` package computes the information criteria (DIC, WAIC, and LOOIC) based on the log-likelihoods obtained from the `log_lik` function.

```{r}
bleval::calc_IC(log_lik_result, 1)
```

### Compute Marginal Likelihood

**Step 1: Specify the `log_prior` function**

The `log_prior` function calculates the log prior density for model parameters. This function takes a named vector of parameter values from a posterior sample and returns the log prior density for each parameter.

```{r}
log_prior <- function(samples_s) {
  lambda <- samples_s[grep("lambda",names(samples_s))]
  beta <- samples_s[grep("beta",names(samples_s))]
  sigma <- samples_s[grep("sigma",names(samples_s))]
  
  log(2) + sum(dgamma(lambda, shape = 0.5, rate = 1, log = TRUE)) +
  sum(dnorm(beta, mean = 0, sd = 1, log = TRUE)) +
  sum(log(truncdist::dtrunc(sigma, "norm", a = 0, b = Inf, mean = 0, sd = 5)))
}
```

**Step 2: Define parameter bounds**

The `log_marg_lik` function requires lower and upper bounds for the model parameters.

```{r}
lb <- c(rep(0, 2), rep(-Inf, 14), rep(0, 2))
ub <- c(rep(Inf, 2), rep(Inf, 14), rep(Inf, 2))

names(lb) <- names(ub) <- colnames(samples)
```

**Step 3: Compute the log marginal likelihood integrating out both latent variables and model parameters**

The `log_marg_lik` function from the `bleval` package computes the log marginal likelihood by integrating out both the latent variables and the model parameters.

```{r}
bleval::log_marg_lik(samples = samples, data = data_list, Ngrid = 5, 
                     lv_mu = lv_mu_list, lv_cov = lv_cov_list, 
                     log_joint_i = log_joint_i, log_prior = log_prior, 
                     lb = lb, ub = ub)
```

The output of this function is the log marginal likelihood, which can be used to compute Bayes factors for model comparison.
