---
title: "Bayesian Evaluation of Multilevel Models Using the bleval Package"
author: "Xiaohui Luo"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document provides a step-by-step guide to evaluating Bayesian multilevel models using the `bleval` R package. We illustrate the workflow with a simulated dataset, performing parameter estimation in JAGS and computing information criteria (DIC, WAIC, LOOIC) and fully marginal likelihood.

## Data Generation

We generate a dataset from a common Gaussian linear mixed model with a random intercept and a random slope.

```{r}
library(MASS) # For multivariate normal distribution
library(dplyr)

set.seed(123)
Nnum <- 500  # Number of level-2 units
Tnum <- 50   # Number of level-1 units
beta <- c(0, 0.3) # Fixed components of random intercept and slope
rho <- 0.3   # Correlation between random intercept and slope
u_cov <- sqrt(1) * sqrt(0.09) * rho
u_sigma <- matrix(c(1, u_cov, u_cov, 0.09), nrow = 2)

bdata <- as.data.frame(mvrnorm(Nnum, mu = beta, Sigma = u_sigma))
names(bdata) <- c("mu_i", "phi_i")
bdata$ID <- c(1:Nnum)

mydata <- data.frame(matrix(NA, nrow = Nnum*Tnum, ncol = 2))
names(mydata) <- c("ID", "t")
mydata$ID <- rep(1:Nnum, each = Tnum)
mydata$t <- rep(1:Tnum, Nnum)

mydata$e_it <- rnorm(n = Nnum*Tnum, mean = 0, sd = 0.1)
mydata$x_it <- rnorm(n = Nnum*Tnum, mean = 0, sd = 1)

mydata <- left_join(mydata, bdata[, c("ID", "phi_i","mu_i")], by = "ID")
mydata$y_it <- mydata$mu_i + mydata$x_it*mydata$phi_i + mydata$e_it

summary(mydata)  # Summary of generated data
```

## Parameter Estimation in JAGS

We define the data-generating Bayesian model in JAGS and estimate parameters using MCMC sampling.

```{r}
library(rjags)

model_string <- "
  model {
    # likelihood ------------------------------
    for (j in 1:Nobs) {
      y[j] ~ dnorm(y_mu[j], y_pre)
      y_mu[j] <- mu[subject[j]] + phi[subject[j]] * x[j]
    }
    for (i in 1:N) {
      mu[i]  <- raneff[i,1]
      phi[i] <- raneff[i,2]
      raneff[i,1:2] ~ dmnorm(beta[1:2], pre[1:2,1:2])
    }
    # priors ---------------------------------
    y_pre ~ dgamma(0.001, 0.001)
    log_pre <- log(y_pre)
    beta[1] ~ dnorm(0, 0.01)
    beta[2] ~ dnorm(0, 0.01)
    beta_0 <- beta[1]
    beta_1 <- beta[2]
    tau_beta_0 ~ dgamma(0.001, 0.001)
    sd_beta_0 <- sqrt(1/tau_beta_0)
    tau_beta_1 ~ dgamma(0.001, 0.001)
    sd_beta_1 <- sqrt(1/tau_beta_1)
    rho ~ dunif(-1, 1)
    sigma[1,1] <- sd_beta_0 * sd_beta_0
    sigma[2,2] <- sd_beta_1 * sd_beta_1
    sigma[1,2] <- sd_beta_0 * rho * sd_beta_1
    sigma[2,1] <- sigma[1,2]
    pre[1:2,1:2] <- inverse(sigma[1:2,1:2])
  }
"

data_list <- list(
  Nobs = nrow(mydata),
  subject = mydata$ID,
  N = length(unique(mydata$ID)),  
  y = mydata$y_it,
  x = mydata$x_it
)

# Initialize the JAGS model
jags_model <- jags.model(textConnection(model_string), data = data_list, n.chains = 4)
# Run the MCMC sampler with a burn-in period of 5000 iterations
update(jags_model, 5000)
# Draw posterior samples from the model
variables <- c("mu", "phi", 
               "beta_0", "beta_1", "y_pre", "tau_beta_0", "tau_beta_1", "rho")
post <- coda.samples(jags_model, variable.names = variables, n.iter = 25000, thin = 10)

# Combine posterior samples from all chains into a single matrix
samps <- do.call(rbind, post)
dim(samps)
# Extract posterior samples for model parameters only (excluding latent variables)
pars_vector <- c("beta_0", "beta_1", "y_pre", "tau_beta_0", "tau_beta_1", "rho")
samps2 <- as.matrix(samps[ ,pars_vector])
dim(samps2)

```

## Model Evaluation with `bleval`

We now evaluate the model using the `bleval` package step by step.

### Installing `bleval`

To use `bleval`, we first need to install the package from GitHub.

```{r}
devtools::install_github("luoxh3/bleval")
```

Then, we can load the package:

```{r}
library(bleval)
```

### Compute Information Criteria

**Step 1: Specify the `log_joint_i` function**

The `log_joint_i` function calculates the log joint density for each unit. This function takes the following inputs:

-   `samples_s`: A named vector of parameter values from a posterior sample.
-   `data`: A list containing the observed data.
-   `i`: The index of the unit.
-   `Ngrid`: The number of quadrature nodes per latent variable.
-   `nodes`: A matrix of latent variable values transformed from the quadrature nodes (output of the `get_quadrature` function).

```{r}
log_joint_i <- function(samples_s, data, i, Ngrid, nodes) {
  # Extract data for unit i
  Nobs <- data$Nobs
  Nnum <- data$N
  Tnum <- Nobs / Nnum
  y_i <- data$y[((i-1)*Tnum+1):(i*Tnum)]
  x_i <- data$x[((i-1)*Tnum+1):(i*Tnum)]
  
  # Expand x and y values for quadrature grid
  x_i_extended_mat <- matrix(rep(x_i, times = Ngrid * Ngrid), 
                             nrow = Ngrid * Ngrid, byrow = TRUE)
  y_i_extended_mat <- matrix(rep(y_i, times = Ngrid * Ngrid), 
                             nrow = Ngrid * Ngrid, byrow = TRUE)
  
  # Compute log conditional likelihood for each unit
  predicted_y <- nodes[, 1] + nodes[, 2] * x_i_extended_mat
  log_con_t_i <- dnorm(y_i_extended_mat, mean = predicted_y, 
                       sd = 1 / sqrt(samples_s[["y_pre"]]), log = TRUE)
  log_con_i <- rowSums(log_con_t_i)
  
  # Compute log prior density for latent variables (random effects)
  sd_mu <- sqrt(1 / samples_s[["tau_beta_0"]])
  sd_phi <- sqrt(1 / samples_s[["tau_beta_1"]])
  mean <- c(samples_s[["beta_0"]], samples_s[["beta_1"]])
  sigma <- matrix(c(sd_mu^2, sd_mu * samples_s[["rho"]] * sd_phi,
                    sd_mu * samples_s[["rho"]] * sd_phi, sd_phi^2), nrow = 2)
  log_raneff_i <- mvtnorm::dmvnorm(nodes, mean, sigma, log = TRUE)
  
  # Return the log joint density for each unit
  log_raneff_i + log_con_i
}
```

**Step 2: Compute the posterior means and covariance matrices of latent variables**

These values are used to adapt the quadrature nodes and weights in the adaptive Gauss-Hermite quadrature.

```{r}
# Create lists to store posterior means and covariance matrices of random effects
raneff_mu_list <- vector("list", Nnum)
raneff_cov_list <- vector("list", Nnum)

for (i in 1:Nnum) {
  # Compute the posterior mean of the random intercept and slope for unit i
  raneff_mu_list[[i]] <- c(
    mean(samps[, paste0("mu[", i, "]")]),
    mean(samps[, paste0("phi[", i, "]")]))
  # Compute the posterior covariance matrix of the random intercept and slope for unit i
  raneff_cov_list[[i]] <- cov(samps[, c(paste0("mu[", i, "]"), 
                                        paste0("phi[", i, "]"))])
}
```

**Step 3: Compute the log marginal likelihood integrating out latent variables**

The `log_marglik` function from the `bleval` package computes the log marginal likelihood by integrating out the latent variables using adaptive Gauss-Hermite quadrature. This function requires the following inputs:

-   `samples`: A matrix or data frame containing the posterior samples of model parameters.
-   `data`: A list of data, including an element 'N' which indicates the number of units
-   `Ngrid`: The number of quadrature nodes for each latent variable.
-   `lv_mu`: A list of posterior means of the latent variables.
-   `lv_cov`: A list of posterior covariance matrices of the latent variables.
-   `log_joint_i`: A user-defined function to compute the log joint density for each unit.

```{r}
log_marglik_result <- bleval::log_marglik(samples = samps2, data = data_list, Ngrid = 9,
                                          lv_mu = raneff_mu_list, lv_cov = raneff_cov_list,
                                          log_joint_i = log_joint_i)
```

The output of this function is a list containing two objects:

-   log_marglik_point: A matrix of log marginal likelihoods for each data point.
-   log_marglik_postmean: A vector of log marginal likelihoods computed using the point estimates (posterior means) of the model parameters.

**Step 4: Compute information criteria**

The `calc_IC` function from the `bleval` package computes the information criteria (DIC, WAIC, and LOOIC) based on the log marginal likelihoods obtained from the `log_marglik` function. This function requires the following inputs:

-   `log_marglik_result`: The output from the `log_marglik` function.
-   `type`: An integer indicating the version of the effective number of parameters (p_D) to use for the DIC calculation. Version 1 is based on the mean of log marginal likelihood values (default), and Version 2 is based on the variance of log marginal likelihood values.

```{r}
bleval::calc_IC(log_marglik_result, 1)
```

The output of this function includes the following information criteria:

-   DIC: Deviance Information Criterion
-   WAIC: Watanabe-Akaike Information Criterion
-   LOOIC: Leave-One-Out Information Criterion

### Compute Fully Marginal Likelihood

**Step 1: Specify the `log_prior` function**

The `log_prior` function calculates the log prior density for model parameters. This function takes a named vector of parameter values from a posterior sample and returns the log prior density for each parameter. The priors used in this example include:

-   Gamma priors for the precision parameters.
-   Normal priors for the fixed effects.
-   A uniform prior for the correlation parameter.

```{r}
log_prior <- function(samples_s) {
  dgamma(samples_s[["y_pre"]], shape = 0.001, rate = 0.001, log = TRUE) +
  dnorm(samples_s[["beta_0"]], mean = 0, sd = sqrt(1/0.01), log = TRUE) +
  dnorm(samples_s[["beta_1"]], mean = 0, sd = sqrt(1/0.01), log = TRUE) +
  dgamma(samples_s[["tau_beta_0"]], shape = 0.001, rate = 0.001, log = TRUE) +
  dgamma(samples_s[["tau_beta_1"]], shape = 0.001, rate = 0.001, log = TRUE) +
  dunif(samples_s[["rho"]], min = -1, max = 1, log = TRUE)
}
```

**Step 2: Define parameter bounds**

The `log_fmarglik` function requires lower and upper bounds for the model parameters. These bounds ensure that the parameter values remain within their valid ranges during the computation of the fully marginal likelihood.

```{r}
lb <- c(rep(-Inf, 2), rep(0, 3), -1)
ub <- c(rep(Inf, 2), rep(Inf, 3), 1)
names(lb) <- pars_vector
names(ub) <- pars_vector # c("beta_0", "beta_1", "y_pre", "tau_beta_0", "tau_beta_1", "rho")
```

**Step 3: Compute the log fully marginal likelihood integrating out both latent variables and model parameters**

The `log_fmarglik` function from the `bleval` package computes the log fully marginal likelihood by integrating out both the latent variables and the model parameters. This function requires the following inputs:

-   `samples`: A matrix or data frame containing the posterior samples of model parameters.
-   `data`: A list of data, including an element 'N' which indicates the number of units
-   `Ngrid`: The number of quadrature nodes for each latent variable.
-   `lv_mu`: A list of posterior means of the latent variables.
-   `lv_cov`: A list of posterior covariance matrices of the latent variables.
-   `log_joint_i`: A user-defined function to compute the log joint density for each unit.
-   `log_prior`: A user-defined function to compute the log prior density for model parameters.
-   `lb`: A named vector of lower bounds for the model parameters.
-   `ub`: A named vector of upper bounds for the model parameters.

```{r}
bleval::log_fmarglik(samples = samps2, data = data_list, Ngrid = 9,
                     lv_mu = raneff_mu_list, lv_cov = raneff_cov_list,
                     log_joint_i = log_joint_i, log_prior = log_prior,
                     lb = lb, ub = ub)
```

The output of this function is the log fully marginal likelihood, which can be used to compute Bayes factors for model comparison.
